### 2020-07-20 code/figure1.Rmd

####################
## Figure 1C data ##
####################
gdc_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>%
  rownames_to_column(var = "sample") %>% ## move path_names to column
  .[match(rownames(tumor_samples), .$sample),]

## umap analysis
set.seed(3296)
gdc.umap <- umap(gdc_path[,2:length(gdc_path)]) ## default n_component is 2
gdc_umap_coors <- data.frame(x_2d = gdc.umap$layout[,1], y_2d = gdc.umap$layout[,2])

####################
## Figure 1E data ##
####################
## To identify clusters using hdbscan, use umap in the third dimension
set.seed(3296)
gdc.umap.3d <- umap(gdc_path[,2:length(gdc_path)], n_components = 3)
gdc.umap.3d <- data.frame(x_3d = gdc.umap.3d$layout[,1], y_3d = gdc.umap.3d$layout[,2], z_3d = gdc.umap.3d$layout[,3]) ## coors of umap
rownames(gdc.umap.3d) <- gdc_path$sample ## insert sample_id in rownames
hdb_cluster_assignment <- hdbscan(gdc.umap.3d, minPts =  114) ## min group size for cluster is 114 (value before cluster number drops significantly)
umap_2d_3d_coors <- data.frame(x_2d = gdc_umap_coors$x_2d, y_2d = gdc_umap_coors$y_2d, x_3d = gdc.umap.3d$x_3d, y_3d = gdc.umap.3d$y_3d, z_3d = gdc.umap.3d$z_3d, clust = as.factor(hdb_cluster_assignment$cluster), member_prob = hdb_cluster_assignment$membership_prob, row.names = rownames(gdc.umap.3d)) ## clust 0 is unclassified cluster

####################
## Figure 1F data ##
####################
## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
umap_2d_3d_coors$knn_clust <- umap_2d_3d_coors$clust
umap_2d_3d_coors$knn_clust[umap_2d_3d_coors$clust==0]<-class::knn(train = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust!=0),],
                                                                  test = umap_2d_3d_coors[,c("x_3d", "y_3d", "z_3d")][which(umap_2d_3d_coors$clust==0),],
                                                                  cl = umap_2d_3d_coors$clust[umap_2d_3d_coors$clust!=0], 
                                                                  k=15)

>>>>

####################
## Figure 1B data ##
####################
gdc_path <- gdc_reactome_path_analysis %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>% ## move path_names to column
  .[match(rownames(tumor_samples), .$sample), ]

gdc_path_mca <- left_join(gdc_path, tumor_samples[, c(1, 3)],
  by = c("sample" = "submitted_sample_id")) %>%
  column_to_rownames(var = "sample")

gdc_path_sample_names <- rownames(gdc_path_mca)
gdc_path_tumor_types <- gdc_path_mca$project_code.x

gdc_path_mca$project_code.x <- NULL

## create a categorical variable from the boolean variable
pathway_list <- make_clean_names(colnames(gdc_path_mca))
pathway_count <- length(pathway_list)
gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
  factor(x, levels = c(0, 1), labels = c("no", "yes"))
}))
colnames(gdc_path_mca) <- pathway_list
gdc_path_mca$project_code <- gdc_path_tumor_types
rownames(gdc_path_mca) <- gdc_path_sample_names

## perform mca
mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)

expl_var <- mca_res$eig[, 2]
expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)

set.seed(4242)
for (k in seq(100)) {
  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
}

randomize <- tibble(
  eigenvalue = 1:length(expl_var),
  expl_var = expl_var,
  perm = colMeans(expl_var_perm),
  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
)

optimal_mca_dimensions <- randomize %>%
  filter(pvalue < 0.05) %>%
  filter(eigenvalue == max(eigenvalue)) %>%
  pull(eigenvalue)

mca_res <- MCA(
  X = gdc_path_mca[, pathway_list],
  ncp = optimal_mca_dimensions, graph = FALSE
)

save(randomize, mca_res, file = "figure1B_data.rda")

####################
## Figure 1C data ##
####################
rng <- RNGseq(max_n_pt * n_iter, 42)
r <- rng[[3649]]

rngtools::setRNG(r)
gdc.umap <- uwot::umap(mca_res$ind$coord, 
                       n_components = 3,
                       n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), # 87
                       n_epochs = 1000,
                       n_threads = 1)
gdc_umap_coors <- as.data.frame(gdc.umap) %>%
  rename(plot_x = V1, plot_y = V2, plot_z = V3)
rownames(gdc_umap_coors) <- rownames(gdc_path_mca)

####################
## Figure 1D data ##
####################
n_iter <- 40
max_n_pt <- 200
score_mca <- matrix(as.numeric(NA), nrow = max_n_pt - 2, ncol = n_iter)

n_tumors <- nrow(mca_res$ind$coord)

cl <- makeCluster(8)
registerDoParallel(cl)

for (i in 3:max_n_pt) {
  print(paste0("MIN POINTS = ", i))
  score_iter <- foreach(1:n_iter,
    .combine = c,
    .export = c("mca_res"),
    .packages = c("dbscan", "uwot"),
    r = rng[(i - 1) * n_iter + 1:n_iter]
  ) %dopar% {
    rngtools::setRNG(r)
    umap_init_mca <- umap(mca_res$ind$coord, 
                          min_dist = 1e-6, 
                          n_components = 3, 
                          n_neighbors = round(sqrt(nrow(mca_res$ind$coord))), 
                          n_epochs = 1000, 
                          n_threads = 1)
    res <- hdbscan(umap_init_mca, minPts = i)
    score_iter_temp <- sum(res$membership_prob < 0.05) / n_tumors
    return(score_iter_temp)
  }
  score_mca[i - 2, ] <- score_iter
}

score <- rowMeans(score_mca)
score_df <- tibble(minPts = 3:max_n_pt, score = score)

sample_df <- score_mca %>%
  as.data.frame() %>%
  as_tibble() %>%
  add_column(minPts = 3:max_n_pt, .before = 1) %>%
  pivot_longer(-minPts) %>%
  group_by(minPts)

optimal_minPts <- score_df %>%
  filter(score == min(score)) %>%
  pull(minPts)

save(score_df, sample_df, optimal_minPts, file = "optimal_minpts_score.rda")

####################
## Figure 1E data ##
####################
rngtools::setRNG(r)

cluster_umap <- uwot::umap(mca_res$ind$coord,
  min_dist = 1e-6,
  n_components = 3,
  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
  n_epochs = 1000,
  n_threads = 1
)
colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
umap_3d_coors <- gdc_umap_coors %>%
  bind_cols(as.data.frame(cluster_umap)) %>%
  mutate(
    clust = as.factor(hdb_cluster_assignment$cluster),
    member_prob = hdb_cluster_assignment$membership_prob,
    row.names = rownames(gdc_umap_coors)
  )
## clust 0 is unclassified cluster

####################
## Figure 1F data ##
####################
## reclassifying clust 0 using our classified cluster as training test and unclassified as test set
umap_3d_coors$clust_knn <- umap_3d_coors$clust
umap_3d_coors$clust_knn[umap_3d_coors$clust == 0] <- class::knn(
  train = umap_3d_coors[which(umap_3d_coors$clust != 0), c("cluster_x", "cluster_y", "cluster_z")],
  test = umap_3d_coors[which(umap_3d_coors$clust == 0), c("cluster_x", "cluster_y", "cluster_z")],
  cl = umap_3d_coors$clust[umap_3d_coors$clust != 0], k = 87
)
